---
title: "Environmental Impact of Housing in England"
author: "Josh Cowley (Newcastle University)"
output:
  pdf_document:
    # https://github.com/Wandmalfarbe/pandoc-latex-template
    template: templates/eisvogel.latex
    number_sections: true
    includes:
      in_header: templates/preamble.tex
---

```{r, echo = FALSE, results = "asis"}
# Load all YAML options stored in separate file
cat(readLines("templates/options.yaml"), sep = "\n")
```

```{r setup, include = FALSE}
LAHS::common_report_setup()
```

\chapter{England Housing Survey (EHS)}

\section{Disclaimer}

Data in this report referred to as the England Housing Survey (EHS) was obtained
from the UK Data Archive (UKDA) under their [End User Licence][ukds-eul] 
agreement.

[ukds-eul]:
  https://ukdataservice.ac.uk/app/uploads/cd137-enduserlicence.pdf

As such, any data collection has been enacted for the purpose of non-profit 
research and further use of the data for commercial use without first obtaining
permission from the data service provider is prohibited.

Furthermore, since these analyses can be interpreted as 'material derived from
the data collections,' any user of this report must also be in agreement to 
these terms.


\clearpage <!------------------------------------------------------------------>


\section{Aims}

- Set out to find population trends and regional differences in said trends.


\clearpage <!------------------------------------------------------------------>


\section{Sampling Methodology}

- Mention the possibility of collection bias and methodology of EHS.
- Need for further analysis of households (not dwellings) for fuel poverty 
  insight.
- Show visualisation of data per year for each region.


\clearpage <!------------------------------------------------------------------>


\section{Dwelling Stock}

\subsection{General}

- EDA into tenure, vacant, alltypex, dampalf.

\subsection{Energy Specific}

- EDA into boiler, dblglaz4, loftinsx, wallinsz, watersys

\subsection{Energy Ratings}

- EDA into sap12, EP_*.
- Correlation with sap12, EPceir12 *and* boiler, dblglaz4, etc.


\clearpage <!------------------------------------------------------------------>


\section{Trends}

- Might get absorbed into previous EDAs or need its own section.


\clearpage <!------------------------------------------------------------------>


\section{Data gaps}

- Renewable technologies in EHS, but available in LAHS.
- Exact CO2 not given in EHS, EPC impact given instead.
- Environmental impact of retrofit.


\clearpage <!------------------------------------------------------------------>


\section{Modelling EPC}

- Aiming to explain EPC using predictors,
- Modelling EPC is only as good as EPC rating quality which is being upgraded 
  soon.


\clearpage <!------------------------------------------------------------------>


\section{Modelling Regional Impact}
- Explain multilevel modelling as way to distinguish population effect and 
  group (region) effect.
