---
title: "Environmental Impact of Housing in England"
author: "Josh Cowley (Newcastle University)"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  pdf_document:
    # https://github.com/Wandmalfarbe/pandoc-latex-template
    template: config/eisvogel.latex
    number_sections: true
    includes:
      in_header: config/preamble.tex
---

```{r setup-yaml, echo = FALSE, results = "asis"}
cat(readLines("config/options.yaml"), sep = "\n")
```

```{r setup, include = FALSE}
library(magrittr) # %>%
LAHS::common_report_setup(fig.asp = 0.55, root.dir = ".")
```


\chapter{English Housing Survey (EHS)}

\section{Disclaimer}

Data in this report referred to as the English Housing Survey (EHS) was obtained
from the UK Data Archive (UKDA) under their [End User Licence][ukds-eul] 
agreement.

As such, any data collection has been enacted for the purpose of non-profit 
research and further use of the data for commercial use without first obtaining
permission from the data service provider is prohibited.

Furthermore, since these analyses can be interpreted as 'material derived from
the data collections,' any user of this report must also be in agreement to 
these terms.


\newpage <!-------------------------------------------------------------------->


\section{Aims}

The aim of this document is threefold:

1. reveal environmental-specific insights into the EHS dataset,

1. model the current EPC rating system based on the factors that can exist in a 
   dwelling,
   
1. understand regional trends in dwelling efficiency while accounting for 
   national trends.

By combining these main aims, we will be able to create an environmental
indicator that serves as a summary to the impact of housing on the climate.

To achieve this, we explore features of the EHS dataset that are pertinent to
the energy efficiency and CO2 emissions of the sampled dwellings. 
Then, we employ regression to better understand the relationships between
the ever-adapting EPC ratings and physical attributes of dwellings such as
insulation and age of boiler installed.

The final aim is realised using a multi-level modelling framework that allows
practitioners to separate population-level effects with group-level effects where
the population in this context is at a national level and the regions, such as
the North East, are groupings.

\section{Sampling Methodology}

The data collected for use in this report are available, with full 
documentation, at the [UK Data Archive (Series 200010)][ukda-series200010].
Documentation compiled by the EHS relay, in detail, how the data was 
collected.

In summary, an initial sample of addresses is pulled from the Postcode Address
File (PAF) using a two-stage random sample that is not covered in any detail
here, for more information see [EHS Technical Reports][ehs-technical].

Participants from this initial sample are invited to take part in an 
\textit{interview} stage survey that is based on \textbf{households}. 
A subset of these participants where permission is granted or the property is
vacant are then subject to a \textit{physical} stage survey.
As such, this 'opt-in' approach may lead to a sample bias that is yet to be
considered.

Note the distinction of terms:

\begin{itemize}
  \item \textbf{household(s)} refer to the group of people (or person) who have
        the accommodation as their only or main residence,
        
  \item \textbf{dwelling(s)} refer to the unit of accommodation which households
        may inhabit.
\end{itemize}

This report concerns itself with dwellings only and does not utilise data from
the initial interview survey. Further work should consider this extension to the
dataset as a means of understanding fuel poverty and the household-dwelling
relationships.

Figure 1.1 shows unequal, yet consistent, sample sizes across regions. This is
expected as the number of houses in each region are not necessarily equal.

\vspace{1.0 em} <!-- NOTE: Make use of space surrounding Figure 1.1 -->
```{r fig1_1, fig.cap = "Unequal sample size across regions"}
LAHS::fig1_1_sample_sizes_by_region()
```

\section{Dwelling Stock}

\subsection{General}

Here we are interested in the general properties of stock sampled in the EHS,
the reader is also advises to view the annual headline report \citep{ehs2020}.
For example, Figure 1.1 in the 2020-21 version shows an increase in private 
renters and a decrease in social renters from the start of the millennium.

Figure 1.2 shows us the proportion of properties that are owned by 
\textit{housing assocations}. Both the North East and North West have
a higher proportion when compared to other regions with no signs of change.
This could be due to a number of reasons such as fewer renters (when compared 
to London) or fewer households being able to get on the property ladder.

```{r fig1_2, fig.cap = "Housing Association market share highest in Northern regions"}
LAHS::fig1_2_percentage_of_ownership()[["Housing association"]]
```

Properties that are vacant are few and far between in the data; all regions
report $96\%$ to $98\%$ of sampled properties are inhabited for the period 
between April 2014 and March 2020.

```{r reftbl_vacant, include = FALSE}
# NOT DISPLAYED. For proof of above statement only.
LAHS::EHS %>%
  dplyr::group_by(.data$gorEHS) %>%
  dplyr::summarise(
    pct_occupied = sprintf("%.1f%%", 100 * mean(.data$vacantx == "Occupied"))
  )
```

Each dwelling is assigned into one of 10 categories based on the type of
dwelling and when it was built. One clear discrepancy in the regions that is 
highlighted by these data is the proportion of flats, defined as a purpose 
built flat or converted flat, as shown in Table 1.1.

```{r tbl1_1}
LAHS::tbl1_1_percentage_of_flats()
```

Also reported in the EHS is if there is any damp present, damp could be a 
symptom of bigger problems. Thankfully, the percentage at which this occurs is
down from 5.18% in 2014-15 to 4.01% in 2019-20 although to understand if
these changes are indicative of an actual decline or just variance in the 
sampling methodology would require further investigation over a larger 
timescale.

```{r reftbl_damp, include = FALSE}
# NOT DISPLAYED. For proof of above statement only.
LAHS::EHS %>%
  dplyr::group_by(.data$YEAR) %>%
  dplyr::summarise(pcT_damp = mean(.data$dampalf == "Damp present"))
```

\subsection{Energy Specific}

<!-- NOTE: Omitted due to time & brevity: dblglaz4, loftinsx, watersys -->

\subsubsection{Space Heating}

Figure 1.3 shows us that there is a clear reliance on gas for systems related to
heating up space with no clear trend emerging from the data. In 
\citet{electricheatingoptions2019}, it is claimed that significant increases
in heating bills and lifetime costs may pose a barrier that would explain this
reliance on gas.

```{r fig1_3, fig.asp = 0.4, fig.cap = "Little to no change in heating systems"}
LAHS::fig1_3_fuelx_trend()
```

\subsubsection{Wall Types}

Wall type data in these data refers to the predominant external wall of a 
dwelling. If a surveyor deems at least 50% of the external wall to be of 
cavity or solid construction, that is recorded, otherwise it is recorded as 
"other".
Similarly, if a surveyor deems at least 50% of the wall to be insulated, 
that is also recorded in the data.
As such, any dwelling can be one of 5 types summarised in Table 1.2


Note that the insulation frequency of cavity walls is much higher than the 
solid walls. As expected due to the difficulties involved with insulating a 
solid wall relative to a cavity type wall.

```{r tbl1_2}
LAHS::tbl1_2_counts_of_wallinsx()
```

\subsubsection{Boiler Type}

For any dwelling that uses central heating, we have information on the type of 
boiler used. If we were to look across the regions we see a similar trend 
occurring that is summarised in Figure 1.4. That is, the volume condensing
boilers in each sample is increasing with time. 

\citet{britishgas2022} claim that condensing boilers these are generally the 
most efficient boilers "on the market" and that "nearly all modern boilers are 
A-rated condensing  boilers" which is a good sign of newer technology that is 
better for the environment is being implemented across all regions.

As with any emerging technology, it is important to remember that a minor 
increase in efficiency may not justify a full boiler replacement and the oldest
or least efficient appliances should be replaced as a priority.

```{r fig1_4, fig.asp = 0.4, fig.cap = "Newer boiler types being adopted, nationwide"}
LAHS::fig1_4_boiler_trend()
```

\subsection{Energy Ratings}

TODO: EDA into sap12, EP_*.

\section{Data Availability}

\subsection{EPC}

Energy and impact ratings are measures that are considered to be indicators 
of efficiency and emissions respectively. However, these ratings are subject to
undergoing changes in an effort to improve their accuracy and utility to
decision makers.
For example, there was a \textit{call for evidence on EPCs} 
\citep{epcaction2018} that aimed to evaluate the current system and improve any
shortcomings.

Hence, in an ideal scenario we would have data on energy used, energy demand 
and CO2 emissions per house, however such an undertaking on a dataset of this 
scale is simply infeasible and unlikely to be improvement on the already
scrutinised EPC measure.

\subsection{Renewable Technologies}

Omitted from the EHS is the mention of renewable technologies (RT).
Such technologies are a significant investment and as such will have a 
substantial impact on a households energy bill and also the energy efficiency
rating.

Datasets such as the Local Authority Housing Data \citep{lahsdata2020}
contain data on the total number of RT(s) a local authority has installed in
its own dwellings that are leased to social renters.
These data also indicate which RT(s) such as heat pumps or biomass boilers but
neglect to reveal the degree of which each is utilised.

\subsection{Retrofit Impact}

As a product of previous improvements, many dwellings are in a condition where
improvements to efficiency or emissions are more 'difficult'. 
That is, a higher economical cost, more resources need or higher skilled labour.
For example, a dwelling with mainly cavity walls is much more likely to be
insulated than a dwelling with solid walls for this reason, see Table 1.2.

Data on the environmental cost of such retrofits are meagre and no such dataset
has been found during the research phase of this project.

To achieve the carbon neutral goals set out by the UK government and others in
the Paris agreement, we must ask is the carbon investment and other pollution
costs worth the offset that they will achieve.
SImilar to various improvements are financially infeasible due to cost, we must
collect more data to rule out improvements that have too high an environmental 
cost.


\newpage <!-------------------------------------------------------------------->


\chapter{Model A} \label{chap:model-a}
\section{Modelling EPC}

An EPC certificate contains various information that is then used to calculate
two typically correlated measures:

1. Energy efficiency rating (EER).

1. Energy impact rating (EIR).

These ratings range from 1 to 100 where higher values indicate dwellings that
are better for the environment. Data is collected using a Standard Assessment
Procedure (SAP) in an attempt to standardise these values across heterogeneous 
properties.

The energy efficiency rating measures the efficiency of the entire building
and is framed to a tenant as a key influence on running costs. That is,
higher values of EER can not only yield benefits for the environment (as less 
energy will be used) but also for the household paying the energy bill.

As shown by the sample certificate provided by the UK Government, the EIR is
more indicative of CO2 emissions \citep{epccertificate2022}.

\vspace{0.6 em}

> The environmental impact rating is a measure of a home’s impact on the 
> environment in terms of carbon dioxide (CO2) emissions. The higher the rating 
> the less impact it has on the environment.

Expectedly, these two ratings tend to be correlated; in our cohort of dwellings
the pearson correlation coefficient is as high as
`r LAHS::display_percent(cor(LAHS::EHS$EPceir12e, LAHS::EHS$sap12), digits = 3)`.

As previously mentioned, the methodology that creates both ratings has been 
under some criticism \citep{epcaction2018} and hence any measure derived from
these data will inherit any limitations.

\subsection{Model Description}

Our initial model comprises of two multivariate linear regressions, one for each
rating described above. The aim of such models is to verify that the variables
that experts would think are important to energy efficiency and emissions are 
then also highly correlated with EER and EIR respectively.

Currently we make use of 6 predictors in our model.

- boiler type,

- double glazing percentage,

- fuel used for space heating,

- degree of loft insulation,

- degree of wall insulation and wall type,

- system used for water heating

We expect all predictors will be statistically significant due to the data 
source, we reason that these specific variables are collected because they are 
seen as important to EPC ratings, and as such this section is a proof of concept
for a greater set of variables.

Greater benefits would be realised from using predictors that are less 
prominent, such as renewable technologies highlighted in the data availability 
section, where parameter estimates would potentially highlight a shortcoming in 
the current rating system.

```{r fit_lms}
fit_sap12 <- 
  stats::lm(
    sap12 ~ boiler + dblglaz4 + fuelx + loftins6 + wallinsz + watersys,
    data = LAHS::EHS
  )

fit_EPceir12e <- 
  stats::lm(
    EPceir12e ~ boiler + dblglaz4 + fuelx + loftins6 + wallinsz + watersys,
    data = LAHS::EHS
  )
```

Table 1.2 shows an estimates for various categories found in the data. One can
calculate an expected EER and EIR by finding the corresponding attributes
and summing the estimates (the intercept is added for all dwellings).

This leads to the intuition that we can compare attributes to each reference 
level and infer its impact on the rating, subject to keeping all other 
attributes fixed.

For example, see that the reference level for boiler type is "Standard" and the 
estimate for "Combination" boiler is 2.6 to 3.3 with a significant P value
($p<0.05$). Thus, we expect that upgrading a house from a standard to a 
combination boiler would increase their EER by approximately 3 points.

```{r tbl2_1}
LAHS::tbl2_1_linear_models(fit_sap12, fit_EPceir12e)
```

\subsection{Discussion}

Table 1.2 shows us that a dwelling with a standard boiler, no double glazing or
insulation that is still reliant on oil systems and an electric water heater
would expect to have a G band EPC rating.
This hypothetical does not meet the current minimum energy efficiency standard
(MEES) allowed for rented properties and such an anti-environment dwelling is 
likely in need of other repairs, and will be potentially vacant.

The impact of different types of boilers is typical with the worst being a 
back boiler, as expected. Interestingly, the impact on efficiency for condensing
and condensing combination boilers appear to be very similar but the latter 
appears to produce less CO2 emissions on average.

Double glazing in many modern homes is a necessity and can be seen as a good
candidate for retrofitting, the data supports this implying a upgrade of 
\textit{all} windows in a house may increase the EER by as many as 7 points.

For space heating there is a single high rating choice, gas fired systems.
There is not significant evidence to distinguish the other options.
As mentioned in an earlier section (see Figure 1.3) electric heating systems
have higher running costs and the efficiency comparison needs to be 
investigated further.
A similar story is told by the water heating estimates, although it should be
noted that a dedicated boiler could include both gas and electric boilers.

Loft insulation appears to increase both EER and EIR as expected but as the 
different between estimates decrease it may be a case of diminishing returns
and warrants further investigation.

Wall insulation is less straightforward to compare since the data has encoded
both insulated and wall type. We have chose the solid uninsulated wall as a 
reference level although modifying walls between solid and cavity is clearly 
an infeasibility and only serves as a comparison on the population scale.

\subsection{Further Work}

As previously mentioned, this model could be improved by increasing the range
of data available, as is the case for almost all models.

One prospect for this model could be decision makers in organisations holding 
dwelling stock as it could serve as a predictive model that highlights any 
properties that have rating less than expected and would benefit from the 
investment of a renewed EPC certificate.

Further work on the current data could be an investigation into interaction
terms, the process of understanding the role of each predictor without the 
restrictive assumption of keeping everything constant. Leading to answers from
questions such as "what is the impact of double glazing in a well insulated
house compared to a less insulated house?".


\newpage <!-------------------------------------------------------------------->


\chapter{Model B}
\section{Modelling Regional Impact}

\subsection{Multilevel Modelling}

Multilevel modelling allows us to distinguish between population effects and 
group level effects \citep{raudenbush2002hierarchical}.
In our context, the population would be all dwellings within England and each
group can be asserted to be the regions. 
Trivially the scope can be decreased, say to a local authority level, or even
increased, say to a worldwide level, depending on the application and available
data.

In our model, we fit no intercept and instead only fit:

1. a population effect for each year that the data spans,

1. a group-level effect for each year that the data spans.

Suppose $n_y$ denotes the number of years our dataset spans; for the EHS data
from 2014 to 2019, $n_y = 6$.
This model will result in $n_y$ population parameters to estimate, 
each summarising the SAP rating, irrespective of regions.
Then, we also have $n_y$ group-level parameters for each defined region that 
summarises how the SAP rating for that region compares to the overall population
on a year-by-year basis.
A formal description of this model is available in Appendix 
\ref{append:mlm-descrip} for the statistically minded reader.

By inferring these parameters we can form an environmental indicator on a 
national level and a more granular indicator for each region that is derived
from the energy efficiency ratings of the English Housing Survey sample.
We focus on efficiency but the methodology can be re-executed for the impact
rating to shift focus from energy use to CO2 emissions.

\subsection{Bayesian Inference}

To fit this model we make use of R and the popular package `brms`
\citep{brms2021}. R version: `r paste(version$major, version$minor, sep = ".")`
and `brms` version: `r utils::packageVersion("brms")`.
The aforementioned software will fit our model in Stan \citep{carpenter2017stan}
that uses Bayesian methods such as Hamiltonian Monte Carlo and a No U-Turn 
sampler, neither of which will be explained in any detail here.

More importantly, these methods allow us to perform Bayesian inference in an
efficient manner. When using the Bayesian paradigm we combine expert knowledge,
known as a \textit{prior}, with what the data tells us, a \textit{likelihood},
to form our \textit{posterior} beliefs.

In this report we use naive priors in the absence of clear expert opinion, 
should such an expert be made available for the benefit of this model, there 
exists many methodologies of prior elicitation that will increase the accuracy
and efficiency of our model, for example see \citet{gosling2018shelf}.

\subsection{Results}

```{r load-fit-mlm, include = FALSE}
# Fit and cached in own script to be used here (see `LAHS/R/scripts`).
fit_mlm <- 
  readRDS(
    file.path(
      rprojroot::find_root("DESCRIPTION"),
      "cache",
      "mlm_regional_epc_run1.rds"
    )
  )
```

We run our MCMC scheme over `r length(fit_mlm$fit@stan_args)` chains for 
`r format(fit_mlm$fit@stan_args[[1]]$iter, big.mark = ",")` iterations 
(`r format(fit_mlm$fit@stan_args[[1]]$warmup, big.mark = ",")` discarded 
as warm-up).
\citet{burkner2017brms} remarks that we need much fewer samples than a 
traditional MCMC scheme due to the higher quality of samples produced,
that is, a higher effective sample size per iteration.

Figure 3.1 shows us the population estimates by year with uncertainty quantified
by the 95% credible intervals of the parameters. Using knowledge of the EPC 
bands, we see that the estimate is representative of a D rating which is not
overly surprising given that the majority of dwellings have a C-D rating.

Positively, we see for the years we have selected the posterior estimate is
increasing indicating an increase in efficiency ratings.

\clearpage <!-- Note: Clear page to force following text below Fig 3.1 -->
```{r fig3_1, fig.asp = 0.5, fig.cap = "Population estimates imply increasing efficiency"}
LAHS::fig3_1_population_effects(fit_mlm)
```

For a region-specific indicator one could combine the estimates from Figure 3.1
with those from Figure 3.2.
However just by investigating the latter Figure, we see that no region's 
estimate(s) are substantially different to 0, implying no region is 
substantially different from the population.
North east credible regions do exceed 0 implying a marginal statistical 
significance but only by a few points at best.

In the author's opinion, small differences on a regional level are likely due 
to the fact that substantial change in housing efficiency tend to come from
\textit{green} government scheme(s) that are available nationwide. Such a claim
warrants further investigation as these models show correlation, not causation.

```{r fig3_2, fig.asp = 0.70, fig.cap = "Region-level estimates not significant"}
LAHS::fig3_2_population_effects(fit_mlm)
```

\subsubsection{Noise Term}

For both models we have assumed an independently and identically normally
distributed error term that is required to allow for variance in the data.
In this model, we summarise our estimate in the same way as the other 
parameters, by looking at the mean and credible region of the samples 
arising from the Markov chain Monte Carlo samples.

```{r warning = FALSE, message = FALSE}
LAHS::tbl3_1_sigma(fit_mlm)
```

The parameter Table 3.1 is trying to estimate is representative of the 
measurement error in the SAP rating and would ideally be minimised. A further
investigation could use this variable as supplementary information to determine
if the accuracy of \textit{Standard Assessment Procedure}.

\subsection{Model Extensibility}

This model, in its current form is closely related to an intercept only 
regression and as such aims to describe data without any ambition to understand
any underlying correlation.

It is straightforward to combine this model described in Chapter 
\ref{chap:model-a} to create a framework that is able to understand the impact 
of various predictors such as boiler type, insulation and others. 
However, unlike before we would be able to understand correlations not only at 
a population level but at a regional level as well.


\newpage <!-------------------------------------------------------------------->


\appendix
\chapter*{Appendix} \addcontentsline{toc}{chapter}{Appendix}
\renewcommand\thesection{\Alph{section}}

\section{Multilevel Model Description} \label{append:mlm-descrip}

TODO: Describe Model B in statistical rigour.

Suppose we let $Y_i$ be the variable we are interested in, such as 
energy efficiency rating, for $i = 1, \dots, n$, where $n$ is the sample size.

In a multivariate linear regression (model A), we 


\newpage <!-------------------------------------------------------------------->


\addcontentsline{toc}{chapter}{Bibliography}


<!-- Links -->

[ukds-eul]:
  https://ukdataservice.ac.uk/app/uploads/cd137-enduserlicence.pdf

[ukda-series200010]:
  https://beta.ukdataservice.ac.uk/datacatalogue/series/series?id=200010
  
[ehs-technical]:
  https://www.gov.uk/government/collections/english-housing-survey-technical-advice
