---
title: "Environmental Impact of Housing in England"
author: "Josh Cowley (Newcastle University)"
output:
  pdf_document:
    # https://github.com/Wandmalfarbe/pandoc-latex-template
    template: config/eisvogel.latex
    number_sections: true
    includes:
      in_header: config/preamble.tex
---

```{r setup-yaml, echo = FALSE, results = "asis"}
cat(readLines("config/options.yaml"), sep = "\n")
```

```{r setup, include = FALSE}
library(magrittr) # %>%
LAHS::common_report_setup(fig.asp = 0.55, root.dir = ".")
```


\chapter{English Housing Survey (EHS)}

\section{Disclaimer}

Data in this report referred to as the English Housing Survey (EHS) was obtained
from the UK Data Archive (UKDA) under their [End User Licence][ukds-eul] 
agreement.

As such, any data collection has been enacted for the purpose of non-profit 
research and further use of the data for commercial use without first obtaining
permission from the data service provider is prohibited.

Furthermore, since these analyses can be interpreted as 'material derived from
the data collections,' any user of this report must also be in agreement to 
these terms.


\clearpage <!------------------------------------------------------------------>


\section{Aims}

The aim of this document is threefold:

1. reveal environmental-specific insights into the EHS dataset,

1. model the current EPC rating system based on the factors that can exist in a 
   dwelling,
   
1. understand regional trends in dwelling efficiency while accounting for 
   national trends.

By combining these main aims, we will be able to create an environmental
indicator that serves as a summary to the impact of housing on the climate.

To achieve this, we explore features of the EHS dataset that are pertinent to
the energy efficiency and CO2 emissions of the sampled dwellings. 
Then, we employ regression to better understand the relationships between
the ever-adapting EPC ratings and physical attributes of dwellings such as
insulation and age of boiler installed.

The final aim is realised using a multi-level modelling framework that allows
practitioners to separate population-level effects with group-level effects where
the population in this context is at a national level and the regions, such as
the North East, are groupings.


\clearpage <!------------------------------------------------------------------>


\section{Sampling Methodology}

The data collected for use in this report are available, with full 
documentation, at the [UK Data Archive (Series 200010)][ukda-series200010].
Documentation compiled by the EHS relay, in detail, how the data was 
collected.

In summary, an initial sample of addresses is pulled from the Postcode Address
File (PAF) using a two-stage random sample that is not covered in any detail
here, for more information see [EHS Technical Reports][ehs-technical].

Participants from this initial sample are invited to take part in an 
\textit{interview} stage survey that is based on \textbf{households}. 
A subset of these participants where permission is granted or the property is
vacant are then subject to a \textit{physical} stage survey.
As such, this 'opt-in' approach may lead to a sample bias that is yet to be
considered.

Note the distinction of terms:

\begin{itemize}
  \item \textbf{household(s)} refer to the group of people (or person) who have
        the accommodation as their only or main residence,
        
  \item \textbf{dwelling(s)} refer to the unit of accommodation which households
        may inhabit.
\end{itemize}

This report concerns itself with dwellings only and does not utilise data from
the initial interview survey. Further work should consider this extension to the
dataset as a means of understanding fuel poverty and the household-dwelling
relationships.

Figure 1.1 shows unequal, yet consistent, sample sizes across regions. This is
expected as the number of houses in each region are not necessarily equal.

```{r fig1_1, fig.cap = "Unequal sample size across regions"}
LAHS::fig1_1_sample_sizes_by_region()
```



\clearpage <!------------------------------------------------------------------>


\section{Dwelling Stock}

\subsection{General}

Here we are interested in the general properties of stock sampled in the EHS,
the reader is also advises to view the annual headline report \citep{ehs2020}.
For example, Figure 1.1 in the 2020-21 version shows an increase in private 
renters and a decrease in social renters from the start of the millennium.

Figure 1.2 shows us the proportion of properties that are owned by 
\textit{housing assocations}. Both the North East and North West have
a higher proportion when compared to other regions with no signs of change.
This could be due to a number of reasons such as fewer renters (when compared 
to London) or fewer households being able to get on the property ladder.

```{r fig1_2, fig.cap = "Housing Association highest in Northern regions"}
LAHS::fig1_2_percentage_of_ownership()[["Housing association"]]
```

Properties that are vacant are few and far between in the data; all regions
report 96% to 98% of sampled properties are inhabited for the period between 
April 2014 and March 2020.

```{r tbl_vacant, include = FALSE}
# NOT DISPLAYED. For proof of above statement only.
LAHS::EHS %>%
  dplyr::group_by(.data$gorEHS) %>%
  dplyr::summarise(
    pct_occupied = sprintf("%.1f%%", 100 * mean(.data$vacantx == "Occupied"))
  )
```

Each dwelling is assigned into one of 10 categories based on the type of
dwelling and when it was built. One clear discrepancy in the regions that is 
highlighted by these data are flats, defined as a purpose built flat or 
converted flat.

```{r tbl1_1}
LAHS::tbl1_1_percentage_of_flats()
```


- EDA into tenure, vacant, alltypex, dampalf.

\subsection{Energy Specific}

- EDA into boiler, dblglaz4, loftinsx, wallinsz, watersys

\subsection{Energy Ratings}

- EDA into sap12, EP_*.
- Correlation with sap12, EPceir12 *and* boiler, dblglaz4, etc.


\clearpage <!------------------------------------------------------------------>


\section{Data gaps}

- Renewable technologies in EHS, but available in LAHS.
- Exact CO2 not given in EHS, EPC impact given instead.
- Environmental impact of retrofit.


\clearpage <!------------------------------------------------------------------>


\section{Modelling EPC}

- Aiming to explain EPC using predictors,
- Modelling EPC is only as good as EPC rating quality which is being upgraded 
  soon.


\clearpage <!------------------------------------------------------------------>


\section{Modelling Regional Impact}

- Explain multilevel modelling as way to distinguish population effect and 
  group (region) effect.


\newpage <!-------------------------------------------------------------------->
\addcontentsline{toc}{chapter}{Bibliography}


<!-- Links -->

[ukds-eul]:
  https://ukdataservice.ac.uk/app/uploads/cd137-enduserlicence.pdf

[ukda-series200010]:
  https://beta.ukdataservice.ac.uk/datacatalogue/series/series?id=200010
  
[ehs-technical]:
  https://www.gov.uk/government/collections/english-housing-survey-technical-advice
